DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-02-13 manuscript/manuscript.Rmd
redraft of downstream anaylisys discussion section

@@

The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
<!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
LLSR: I changed loosing to blurring. But I will elaborate more on it}
Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
\textcolor{red}{LLSR: Example}-->

@@

## Effects on downstream analyses

The study of phenomena dependent on the timing of species diversification events, such as macroevolutionary processes, is affected by the usage of alternative chronograms that vary in topology [@rabosky2015no; @title2016macrophylogenies].
<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->

On the other hand, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008].
Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.

Adding branch lengths sampled from a birth-death model in the absence of genetic data
has been found to improve insight in phylogeny-based analyses.

is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.

Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic [@cusimano2012new].

@thomas2013pastis hypothesize that results of diversification analyses will be biased towards the birth-death model used to simulate branch lengths;
and note that "the effects of missing species placement or polytomy resolution are less clear for other phylogeny-based analyses (e.g. correlates of diversification, modelling trait evolution, community phylogenetics), and future work should test how the treatment of missing species influences both parameter estimation and type I and II errors"

Notably, risks come with this practice.

Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.

>>>>

While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008] and diversiifcation [@], it is important to note possible sources of problems when reusing chronogram data the way DateLife does.

@@

The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
<!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
LLSR: I changed loosing to blurring. But I will elaborate more on it}
Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
\textcolor{red}{LLSR: Example}-->

@@

## Effects of phylogenetic sampling on downstream analyses

Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating]. 
Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many taxa.
Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
Completing a phylogeny using birth-death politomy resolvers and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These phylogenies provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns [@]. 
Analysis of empirical phylogenies suggest that introducing stochastically resolved divergence times creates a spurious pattern of early burst of diversification.

Taxonomy based politomy resolvers also introduce topological differences.
The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group. 
For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization. We note that the same could apply for geographical and morphological dependent diversification analysis. Hence, we suggest that these types of randomly resolved phylogenies can work as null or neutral models, representing the case of diversification independent of traits and geographical scenario.

Notably, risks come with the practice of resolving politomies, even if using a tested model of diversification.
Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods. 
We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
